---
title: "Lecture 6. Data Validation"
author: "Zechariah Meunier"
date: "`r Sys.Date()`"
output: html_document
---

# Definitions
**Data validation** - process of ensuring that data are accurate (i.e., have correct values) and precise (i.e., have consistent values)

**Quality assurance (QA)** - systematic prevention of defects in a product or service so that they meet specified requirements

- proactive approach
- requires audit of the process or **input**

**Quality control (QC)** - identification and rectification of defects in a product or service so that they meet specified requirements

- reactive approach
- requires audit of the product or **output**

# Quality assurance

### NOAA Data Example
We'll begin with a simple example of a dataset from the National Oceanic and Atmospheric Administration (NOAA). These data are the bimonthly values for the Multivariate El Niño Southern Oscillation Index (https://psl.noaa.gov/enso/mei/) or MEI for short.

First, let's load the input data. These data were last updated on March 10, 2025.
```{r}
MEI_raw <- read.delim("MEI_raw.txt")
MEI <- read.table("MEI_cleaned.txt", header = TRUE)
```

Now take a look at the MEI_raw vs. MEI_cleaned files. In this case, it was easiest to work with the data in a text editor, but larger files might need to be manipulated in R.
```{r}
View(MEI_raw)
View(MEI)
```

To check how R stores the data in the table and columns, use the `class` function.
```{r}
class(MEI)
class(MEI$YEAR)
class(MEI$DJ)
```

Now let's get a sense of the spread of the data using the function `summary`.
```{r}
summary(MEI)
```

Another nice way of summarizing data is to use the `boxplot` function.
```{r}
boxplot(MEI)
```

As you can see, the year column is distorting the y-axis, so let's remove it using a trick by indexing into the columns and removing the first column.
```{r}
boxplot(MEI[,-1])
```

However, there's still some issues with major outliers. So let's remove the last row also. If you don't know how many rows are present, use the `nrow` function. Similarly, you can use `ncol` to determine the number of columns present.
```{r}
boxplot(MEI[-nrow(MEI),-1])
```
 
To view the top few rows of the data frame, use the `head` function. Similarly, use the `tail` function to view the bottom few rows.
```{r}
head(MEI)
tail(MEI)
```
As you can see, there's an issue with the last row having -999.000 as a placeholder instead of `NA`. There are several ways to convert these values, some of which are more work than others.

We could manually replace the values in a vector using `replace` and fix it in 9 lines of code.
```{r}
MEI$FM <- replace(MEI$FM, MEI$FM == -999, NA)
MEI$MA <- replace(MEI$MA, MEI$MA == -999, NA)
MEI$AM <- replace(MEI$AM, MEI$AM == -999, NA)
MEI$MJ <- replace(MEI$MJ, MEI$MJ == -999, NA)
MEI$JJ <- replace(MEI$JJ, MEI$JJ == -999, NA)
MEI$JA <- replace(MEI$JA, MEI$JA == -999, NA)
MEI$AS <- replace(MEI$AS, MEI$AS == -999, NA)
MEI$SO <- replace(MEI$SO, MEI$SO == -999, NA)
MEI$ON <- replace(MEI$ON, MEI$ON == -999, NA)
MEI$ND <- replace(MEI$ND, MEI$ND == -999, NA)
tail(MEI)
```

We could use a for-loop and fix it in 4 lines of code.
```{r}
MEI <- read.table("MEI_cleaned.txt", header = TRUE)

for (i in 1:ncol(MEI)) {
  col <- colnames(MEI)[i]
  MEI[[col]] <- replace(MEI[[col]], MEI[[col]] == -999, NA)
}

tail(MEI)
```

We could use the `lapply` function and fix it in 1 line of code.
```{r}
MEI <- read.table("MEI_cleaned.txt", header = TRUE)

MEI[] <- lapply(MEI, function(x) replace(x, x == -999, NA))

tail(MEI)
```


# Quality control